Simple repression draft

### 2020-10-14 doc/resubmission/section_02_means.tex
update to latest version of main text

\fig{fig1:means_cartoons} depicts a smorgasbord of mathematicized cartoons for
simple repression using both thermodynamic and kinetic models that have appeared
in previous literature. For each cartoon, we calculate the fold-change in mean
gene expression as predicted by that model, deferring some algebraic details to
Appendix~\ref{sec:non_bursty}. What we will find is that all cartoons collapse
to a single master curve as shown in \fig{fig1:means_cartoons}(D), which
contains just two parameters. We label the parameters $\Delta F_R$, an effective
free energy parametrizing the repressor-DNA interaction, and $\rho$, which
subsumes all details of transcription in the absence of repressors. We will
offer some intuition for why this master curve exists and discuss why at the
level of the mean expression, we are unable to discriminate ``right'' from
``wrong'' cartoons given only measurements of fold-changes in expression.

@@

Again, we consign all details of the derivation to Appendix~\ref{sec:non_bursty}.
Here we just highlight the general findings for all five kinetic models. As
already shown in Figure~\ref{fig1:means_cartoons}(C) and (D), all the kinetic
models explored can be collapsed onto the master curve. Given that the
repressor-bound state only connects to the rest of the promoter dynamics via its
binding and unbinding rates, $k_R^+$ and $k_R^-$ respectively, all models can
effectively be separated into two categories: a single repressor-bound state,
and all other promoter states with different levels of coarse graining. This
structure then guarantees that, at steady-state, detailed balance between these
two groups is satisfied. What this implies is that the steady-state distribution
of each of the non-repressor states has the same functional form with or without
the repressor, allowing us to write the fold-change as a product of the ratio of
the binding and unbinding rates of the promoter, and the promoter details. This
is 
\mrm{This is based in one of the comments from reviewer \# 2. It is empirically
true for all models we explored, but I don't know how to prove that it is a 
general statement for any unimaginable complex model.}
\begin{align}
\text{fold-change} &= \left( 1 + \frac{k_R^+}{k_R^-} \rho \right)^{-1},\\
&= (1 + \exp(-\Delta F_R + \log(\rho) ))^{-1},
\end{align}
where $\Delta F_R \equiv -\log(k_R^+/k_R^-)$, and the functional forms of $\rho$
for each model change as shown in Figure~\ref{fig1:means_cartoons}(C) (See
Appendix~\ref{sec:non_bursty} for further details).

>>>>

Figure~\ref{fig1:means_cartoons}(B) and (C) depicts a smorgasbord of
mathematicized cartoons for simple repression using both thermodynamic and
kinetic models, respectively, that have appeared in previous literature. For
each cartoon, we calculate the fold-change in mean gene expression as predicted
by that model, deferring most algebraic details to
Appendix~\ref{sec:non_bursty}. What we will find is that for all cartoons the
fold-change can be written as a Fermi function 
\begin{equation}
 \text{fold-change} = \left( 1 + \exp(-\Delta F_R + \log(\rho))  \right)^{-1},
 \label{eq:deltaFR_eq_noneq_equiv}
\end{equation}
where the effective free energy contains two terms: the parameters $\Delta F_R$,
an effective free energy parametrizing the repressor-DNA interaction, and
$\rho$, a term derived from the level of coarse-graining used to model all
repressor-free states. In other words, the effective free energy of the Fermi
function can be written as the additive effect of the regulation given by the
repressor via $\Delta F_R$, and the kinetic scheme used to describe the steps
that lead to a transcriptional event via $\log(\rho)$ (See
Figure~\ref{fig1:means_cartoons}(D), left panel). This implies that all models
collapse to a single master curve as shown in
Figure~\ref{fig1:means_cartoons}(D). We will offer some intuition for why this
master curve exists and discuss why at the level of the mean expression, we are
unable to discriminate ``right'' from ``wrong'' cartoons given only measurements
of fold-changes in expression.

@@

Again, we consign all details of the derivation to
Appendix~\ref{sec:non_bursty}. Here we just highlight the general findings for
all five kinetic models. As already shown in Figure~\ref{fig1:means_cartoons}(C)
and (D), all the kinetic models explored can be collapsed onto the master curve.
Given that the repressor-bound state only connects to the rest of the promoter
dynamics via its binding and unbinding rates, $k_R^+$ and $k_R^-$ respectively,
all models can effectively be separated into two categories: a single
repressor-bound state, and all other promoter states with different levels of
coarse graining. This structure then guarantees that, at steady-state, detailed
balance between these two groups is satisfied. What this implies is that the
steady-state distribution of each of the non-repressor states has the same
functional form with or without the repressor, allowing us to write the
fold-change as a product of the ratio of the binding and unbinding rates of the
promoter, and the promoter details. This is \mrm{This is based in one of the
comments from reviewer \# 2. It is empirically true for all models we explored,
but I don't know how to prove that it is a general statement for any
unimaginable complex model.}
\begin{align}
\text{fold-change} &= \left( 1 + \frac{k_R^+}{k_R^-} \rho \right)^{-1},\\
&= (1 + \exp(-\Delta F_R + \log(\rho) ))^{-1},
\end{align}
where $\Delta F_R \equiv -\log(k_R^+/k_R^-)$, and the functional forms of $\rho$
for each model change as shown in Figure~\ref{fig1:means_cartoons}(C). Another
intuitive way two think about these to terms is as follows: in all kinetic
models shown in Figure~\ref{fig1:means_cartoons}(C) the repressor-bound state
can only be reached from a single repressor-free state. The ratio of these two
states --repressor-bound and adjacent repressor-free state-- must remain the
same for all models, regardless of the details included in other promoter states
if $\Delta F_R$ represents an effective free energy of the repressor binding the
DNA operator. The presence of other states then draws probability density from
the promoter being in either of these two states, making the ratio between the
repressor-bound state and \textit{all} repressor-free states different. The log
difference in this ratio is given by $\log(\rho)$. Since model 1 and model 5 of
Figure~\ref{fig1:means_cartoons}(C) consist of a single repressor-free state,
$\rho$ is then necessarily 1 (See Appendix~\ref{sec:non_bursty} for further
details).

### 2020-10-14 doc/resubmission/section_04_bayesian_inference.tex
update to latest version of main text

In this section we aim to reconcile the thermodynamic and kinetic perspectives
that have been brought to bear on the repressor-operator model. From
Section~\ref{section_02_means}, this reconciliation depends on a key
quantitative question as codified by Eq.~\ref{eq:deltaFR_eq_noneq_equiv}: does
the free energy of repressor binding, as described in the thermodynamic models
and indirectly inferred from gene expression measurements, agree with the
corresponding values of repressor binding and unbinding rates in the kinetic
picture, measured or inferred more directly? In this section we tackle the
statistical inference problem of inferring these repressor rates from
single-cell mRNA counts data. But before we can turn to the full case of simple
repression, we must choose an appropriate model of the constitutive promoter and
infer the parameter values in that model. This is the problem we address first.

@@

From consideration of Fano factors in the previous section, we suspect that
model 5 in Figure~\ref{fig2:constit_cartoons}(A), a one-state bursty model of
constitutive promoters, achieves the right balance of complexity and simplicity,
by allowing both Fano factor $\nu>1$, but also by remedying, by design, the
problems of parameter degeneracy that model 4 in
Figure~\ref{fig2:constit_cartoons}(A) suffered~\cite{Razo-Mejia2020}. We will
test this thoroughly on single-cell mRNA counts for different unregulated
promoters from Jones et.\ al.~\cite{Jones2014}, but first we consider It will be
instructive, however, to first consider the Poisson promoter, model 1 in
Figure~\ref{fig2:constit_cartoons}. As we alluded to earlier, since the Poisson
distribution has a Fano factor $\nu$ strictly equal to 1, and all of the
observed data in Figure~\ref{fig2:constit_cartoons}(B) has Fano factor $\nu>1$,
we might already suspect that this model is incapable of fitting the data. We
will verify that this is in fact the case. Using the same argument we can
immediately rule out models 2 and 3 from Figure~\ref{fig2:constit_cartoons}(A).
These models have Fano factors $\nu\le 1$ meaning they are underdispersed
relative to the Poisson distribution. We will also not explicitly consider model
4 from~\fig{fig2:constit_cartoons}(A) since it was already thoroughly analyzed
in~\cite{Razo-Mejia2020}, and since model 5 can be viewed as a special case of
it.

\mrm{Griffin also feels this paragraph should go. Too many times reminding the 
reader what we are up to. I think he is right.}
Our objective for this section will then be to assess whether or not model 5 is
quantitatively able to reproduce experimental data. In other words, if our claim
is that the level of coarse graining in this model is capable of capturing the
relevant features of the data, then we should be able to find values for the
model parameters that can match theoretical predictions with single-molecule
mRNA count distributions. A natural language for this parameter inference
problem is that of Bayesian probability. We will then build a Bayesian inference
pipeline to fit the model parameters to data. To gain intuition on how this
analysis is done we will begin with the ``wrong'' model 1 in
Figure~\ref{fig2:constit_cartoons}(A). We will use the full dataset of
single-cell mRNA counts from~\cite{Jones2014} used in
Figure~\ref{fig2:constit_cartoons}(B).

>>>>

In this section of the paper, we continue our program of providing one complete
description of the entire broad sweep of studies that have been made in the
context of the repressor-operator model, dating all the way back to the original
work of Jacob and Monod and including the visionary quantitative work of
M\"{u}ller-Hill and collaborators~\cite{Oehler1990} and up to more recent
studies \cite{Garcia2011a}. In addition, the aim is to reconcile the
thermodynamic and kinetic perspectives that have been brought to bear on this
problem. From Section~\ref{section_02_means}, this reconciliation depends on a
key quantitative question as codified by Eq.~\ref{eq:deltaFR_eq_noneq_equiv}:
does the free energy of repressor binding, as described in the thermodynamic
models and indirectly inferred from gene expression measurements, agree with the
corresponding values of repressor binding and unbinding rates in the kinetic
picture, measured or inferred more directly? In this section we tackle the
statistical inference problem of inferring these repressor rates from
single-cell mRNA counts data. But before we can turn to the full case of simple
repression, we must choose an appropriate model of the constitutive promoter and
infer the parameter values in that model. This is the problem we address first.

@@

From consideration of Fano factors in the previous section, we suspect that
model 5 in Figure~\ref{fig2:constit_cartoons}(A), a one-state bursty model of
constitutive promoters, achieves the right balance of complexity and simplicity,
by allowing both Fano factor $\nu>1$, but also by remedying, by design, the
problems of parameter degeneracy that model 4 in
Figure~\ref{fig2:constit_cartoons}(A) suffered~\cite{Razo-Mejia2020}. Does this
stand up to closer scrutiny, namely, comparison to full mRNA distributions
rather than simply their moments? We will test this thoroughly on single-cell
mRNA counts for different unregulated promoters from Jones et.\
al.~\cite{Jones2014}.

It will be instructive, however, to first consider the Poisson promoter, model 1
in Figure~\ref{fig2:constit_cartoons}. As we alluded to earlier, since the
Poisson distribution has a Fano factor $\nu$ strictly equal to 1, and all of the
observed data in Figure~\ref{fig2:constit_cartoons}(B) has Fano factor $\nu>1$,
we might already suspect that this model is incapable of fitting the data. We
will verify that this is in fact the case. Using the same argument we can
immediately rule out models 2 and 3 from Figure~\ref{fig2:constit_cartoons}(A).
These models have Fano factors $\nu\le 1$ meaning they are underdispersed
relative to the Poisson distribution. We will also not explicitly consider model
4 from~\fig{fig2:constit_cartoons}(A) since it was already thoroughly analyzed
in~\cite{Razo-Mejia2020}, and since model 5 can be viewed as a special case of
it.

Our objective for this section will then be to assess whether or not model 5 is
quantitatively able to reproduce experimental data. In other words, if our claim
is that the level of coarse graining in this model is capable of capturing the
relevant features of the data, then we should be able to find values for the
model parameters that can match theoretical predictions with single-molecule
mRNA count distributions. A natural language for this parameter inference
problem is that of Bayesian probability. We will then build a Bayesian inference
pipeline to fit the model parameters to data. To gain intuition on how this
analysis is done we will begin with the ``wrong'' model 1 in
Figure~\ref{fig2:constit_cartoons}(A). We will use the full dataset of
single-cell mRNA counts from~\cite{Jones2014} used in
Figure~\ref{fig2:constit_cartoons}(B).
